Equations & R Code

### 2017-02-20 manuscript/coexistence-stability_ms.tex
updates equations; modifies all use of variability in text as either ecosystem or environment

\renewcommand\linenumberfont{\normalfont\tiny\sffamily\color{gray}}
\newcommand{\new}{\textcolor{blue}}

@@

MacArthur (1955), Elton (1958), and even Darwin (Turnbull et al. 2013)
recognized the potential for compensatory dynamics among species to
stabilize ecosystem functioning in fluctuating environments. This idea
underlies the ``insurance hypothesis'' (Yachi \& Loreau 1999), which
suggests ecosystem variability decreases with diversity because species
respond dissimilarly to environmental variation, broadening the range of
conditions under which the community maintains function (Loreau 2010). A
variety of theoretical models all predict a negative relationship
between species richness and ecosystem variability (Lehman \& Tilman
2000; Ives \& Hughes 2002; Loreau \& de Mazancourt 2013), and
experimental tests tend to support such a prediction (Tilman et al.
2006; Hector et al. 2010).

However, the ability of biodiversity-ecosystem functioning (BEF)
experiments to accurately represent real-world dynamics is debated
(Eisenhauer et al. 2016; Wardle 2016). Much of the debate centers on the
fact that BEF experimental protocols do not allow species gains to
offset species losses. Theoretical work on diversity-variability
relationships typically suffers from the same limitation: it recognizes
the role of environmental variability in driving population fluctuations
which destabilize ecosystems, but ignores the potential for
environmental variability to promote species richness and thereby help
stabilize ecosystems (Loreau 2010, but see Chesson et al. 2001).

@@

The opposing effects of environmental variability on ecosystem
variability might explain the mixed results from empirical studies on
the diversity-variability relationship. Observational tests of the
diversity-variability relationship, which require sampling across
natural diversity gradients, have yielded negative (Hautier et al.
2014), neutral (Valone \& Hoffman 2003; Cusson et al. 2015), and
positive (Sasaki \& Lauenroth 2011) relationships. In a meta-analysis of
diversity-variability relationships based on observational studies in
terrestrial ecosystems, Jiang \& Pu (2009) found no significant evidence
for an effect of species richness on ecosystem variability, perhaps
because environmental variability varies across natural diversity
gradients, affecting both richness and ecosystem variability. The
idiosyncratic results of these empirical studies contrast with the
consistent conclusions from theoretical work that ignores feedbacks
between variability and richness.

The gap between theoretical expectations and empirical results of
richness-variability relationships might reflect the divergence of
theory developed to explain species coexistence and theory developed to
explain diversity and ecosystem variability. One reason these two
disciplines have diverged is because they have focused on different
questions. Diversity-variability studies typically ask how ecosystem
variability responds to different levels of species richness at a given
level of environmental variability (reviewed in Kinzig et al. 2001;
Loreau 2010), whereas coexistence studies ask how species richness
responds to different levels of environmental variability (Chesson \&
Warner 1981).

@@

We developed a semi-discrete consumer-resource model that allows many
species to coexist on one resource by either the storage effect or
relative nonlinearity. In our model, the consumer can be in one of
two-states: a dormant state \(D\) and a live state \(N\). The dormant
state could represent, for example, the seed bank of an annual plant
\new{or root biomass of a perennial plant}. Transitions between \(N\)
and \(D\) occur at discrete intervals \(\tau\) with continuous-time
consumer-resource dynamics between discrete transitions. Thus, our model
is formulated as ``pulsed differential equations'' (Pachepsky et al.
2008; Mailleret \& Lemesle 2009; Mordecai et al. 2016). For clarity we
refer to \(\tau\) as years and the growing time between years as seasons
with daily (\(t\)) time steps.

@@

\begin{align}
N_i(\tau^+) = \gamma_{i,\tau} D_i(\tau^+),
\end{align}\vspace{-3em}

\noindent \new{where \emph{i} indexes each species and $\tau^+$ denotes the beginning of growing season $\tau$.}
\new{Live biomass at the start of the growing season ($N_i(\tau^+)$) then serves as the initial conditions for continuous-time}
consumer-resource dynamics that are modeled as two differential
equations: \vspace{-3em}

@@

\new{At the end of the growing season, a fraction ($\alpha_i$) of live biomass is stored as dormant biomass and a fraction of dormant biomass survives ($\eta_i$) to the next growing season, giving the following equation:}
\vspace{-3em}

\begin{align}
D_i{\tau + 1} = \left[\alpha_i N_i(\tau^-) + D_i(\tau^-) \right](1 - \eta_i)
\end{align}\vspace{-3em}

\noindent \new{ where $\tau^-$ denotes the end of growing season $\tau$.
We assume remaining live biomass ($N(\tau^-)(1-\alpha)$) dies (i.e., this is not a closed system where all biomass must be in either $N$ or $D$ states).}

We assume the resource pool is not replenished within a growing season.
Resource replenishment occurs between growing seasons, and the resource
pool (\emph{R}) at the start of the growing season is
\(R(\tau^+) = R^+\), where \(R^+\) is a random resource pulse drawn from
a log-normal distribution with mean \(\mu(R^+)\) and standard deviation
\(\sigma(R^+)\).
\new{Taken all together, we can combine equations 1 and 4 to define the discrete transitions between live and dormant biomass at the end of a growing season.
Thus, the initial conditions for each state ($N, D, R$) at the beginning of growing season $\tau+1$ are:}
\vspace{-3em}

\begin{align}
  D_{i}(\tau^+ + 1) &= (1-\gamma_{i,\tau}) \left[\alpha_i N_{i}(\tau^-) + D_{i}(\tau^-) \right] (1-\eta_i) \\
  N_{i}(\tau^+ + 1) &= \gamma_{i,\tau} \left[\alpha_i N_{i}(\tau^-) + D_{i}(\tau^-) \right] (1-\eta_i) \\
  R(\tau^+ + 1) &= \text{lognormal}\left(\mu(R^+), \sigma(R^+) \right),
\end{align}\vspace{-3em}

\noindent \new{where, as above, $\tau^-$ denotes the end of growing season $\tau$ and $\tau^+ + 1$ denotes the beginning of growing season $\tau+1$.}
Model parameters and notation are described in Table 1.

@@

When we allowed the variance of the environment to determine which of
four initial species coexisted, similar to a study across a natural
diversity gradient, we found a positive relationship between richness
and variability (Fig. 2A,C). This was true for the storage effect, where
coexistence is maintained by fluctuating dormant-to-live transition
rates (\(\gamma\)), and for relative nonlinearity, where coexistence is
maintained by annual resource pulses. The relationship is driven by the
fact that increasing environmental variability increases the strength of
both coexistence mechanisms (Fig. S1-2). More variable conditions
promoted species richness, creating a positive relationship between
diversity and variability.

When we performed species removals but held environmental variability at
a level that allows coexistence of all four species, similar to a
biodiversity-ecosystem functioning experiment, we found a negative
diversity-variability relationship (Fig. 2B,D). Scatter around the
relationship was small for the storage effect because all species have
similar temporal variances. Regardless of species identity, the presence
of more species always stabilized ecosystem functioning through
portfolio effects. In contrast, scatter around the relationship was
larger for relative nonlinearity (Fig. 2D) because species with
different resource uptake curves had different population variances.
Depending on which species were present, two-species communities were
sometimes less variable than three-species communities.

@@

The dampening effect of fluctuation-dependent coexistence on increasing
environmental variability depends on the specific traits (parameter
values) of the species in the regional pool. Moderate competition makes
it more difficult for new species to enter the local community, but once
they do enter, variability is similar between communities with low and
moderate competition (Fig. 3; compare top and bottom panels). Moderate
competition does increase the rate at which ecosystem \emph{CV}
increases with environmental variance (Fig. S1-3) because the abundance
of inferior competitors is reduced and they do not influence ecosystem
\emph{CV} as much as when competition is low. The correlation of
species' environmental responses also mediates the relationship between
environmental variance, species richness, and ecosystem \emph{CV}: lower
correlations make it easier for new species to enter the community and
contribute to porfolio effects (Fig. 3).

@@

Theory developed for biodiversity-ecosystem function experiments
emphasizes that increases in species richness should reduce ecosystem
variability. Consistent with theoretical expectations from models and
with results from biodiversity-ecosystem functioning experiments, in
which species coexistence is maintained by fluctuation-independent
mechanisms, our model of fluctuation-dependent species coexistence (also
see Chesson et al. 2001) produced a negative diversity-variability
relationship (Fig. 2B,D). This agreement is encouraging because species
almost certainly coexist by some combination of fluctuation-independent
(e.g., resource partitioning) and fluctuation-dependent mechanisms. By
extending theory to communities where species richness is explicitly
maintained by temporal variability, we have gained confidence that
experimental findings are generalizable to many communities. In local
settings where environmental variability is relatively homogeneous,
reductions in the number of species should increase the variability of
ecosystem functioning, regardless of how coexistence is maintained.

@@

Our results may explain why deviations from the negative
diversity-variability relationship often come from observational studies
(Jiang \& Pu 2009). Observational studies must rely on natural diversity
gradients, which do not control for differences in environmental
variability among sites. If species richness depends on environmental
variability, it is entirely possible to observe positive
diversity-variability relationships. For example, Sasaki and Lauenroth
(2011) found a positive relationship between species richness and the
temporal variability of plant abundance in a semi-arid grassland. Their
data came from a six sites that were 6 km apart. While Sasaki and
Lauenroth explained their results in terms of dominant species' effects,
it is also possible that each site experienced sufficiently different
levels of environmental variability to influence species coexistence.
DeClerck et al. (2006) also found a positive diversity-variability when
sampling conifer richness and the variability of productivity across a
large spatial gradient in the Sierra Nevada.

While our modeling results show that fluctuation-dependent coexistence
can create positive diversity-variability relationships, whether such
trends are detected will depend on the particular traits of the species
in the community and the relative influence of fluctuation-dependent and
fluctuation-independent coexistence mechanisms. Thus, our results may
also help explain observational studies where no relationship between
diversity and variability is detected. For example, Cusson et al. (2015)
found no relationship between species richness and variability of
abundances in several marine macro-benthic ecosystems. Many of their
focal ecosystems were from highly variable intertidal environments. If
coexistence was at least in part determined by environmental
fluctuations, then the confounding effect of variability and species
richness could offset or overwhlem any effect of species richness on
variability. Previous theoretical work showed how environmental
variation can mask the effect of species diversity on ecosystem
productivity when sampling across sites (Loreau 1998). Our mechanistic
model extends that conclusion to ecosystem variability.

>>>>

\renewcommand\linenumberfont{\normalfont\tiny\sffamily\color{gray}}

\definecolor{blue}{rgb}{0,0,0.7} \newcommand{\new}{\textcolor{blue}}

@@

MacArthur (1955), Elton (1958), and even Darwin (Turnbull et al. 2013)
recognized the potential for compensatory dynamics among species to
stabilize ecosystem functioning in fluctuating environments. This idea
underlies the ``insurance hypothesis'' (Yachi \& Loreau 1999), which
suggests
\new{the variation of ecosystem biomass through time (ecosytem variability)}
decreases with diversity because species respond dissimilarly to
environmental variation, broadening the range of conditions under which
the community maintains function (Loreau 2010). A variety of theoretical
models all predict a negative relationship between species richness and
ecosystem variability (Lehman \& Tilman 2000; Ives \& Hughes 2002;
Loreau \& de Mazancourt 2013), and experimental tests tend to support
such a prediction (Tilman et al. 2006; Hector et al. 2010).

However, the ability of biodiversity-ecosystem functioning (BEF)
experiments to accurately represent real-world dynamics is debated
(Eisenhauer et al. 2016; Wardle 2016). Much of the debate centers on the
fact that BEF experimental protocols do not allow species gains to
offset species losses. Theoretical work on
diversity-\new{ecosystem variability} relationships typically suffers
from the same limitation: it recognizes the role of environmental
variability in driving population fluctuations which destabilize
ecosystems, but ignores the potential for environmental variability to
promote species richness and thereby help stabilize ecosystems (Loreau
2010, but see Chesson et al. 2001).

@@

The opposing effects of environmental variability on ecosystem
variability might explain the mixed results from empirical studies on
the diversity-\new{ecosystem variability} relationship. Observational
tests of the diversity-\new{ecosystem variability} relationship, which
require sampling across natural diversity gradients, have yielded
negative (Hautier et al. 2014), neutral (Valone \& Hoffman 2003; Cusson
et al. 2015), and positive (Sasaki \& Lauenroth 2011) relationships. In
a meta-analysis of diversity-\new{ecosystem variability} relationships
based on observational studies in terrestrial ecosystems, Jiang \& Pu
(2009) found no significant evidence for an effect of species richness
on ecosystem variability, perhaps because environmental variability
varies across natural diversity gradients, affecting both richness and
ecosystem variability. The idiosyncratic results of these empirical
studies contrast with the consistent conclusions from theoretical work
that ignores feedbacks between variability and richness.

The gap between theoretical expectations and empirical results of
diversity-\new{ecosystem variability} relationships might reflect the
divergence of theory developed to explain species coexistence and theory
developed to explain diversity and ecosystem variability. One reason
these two disciplines have diverged is because they have focused on
different questions. Diversity-\new{ecosystem variability} studies
typically ask how ecosystem variability responds to different levels of
species richness at a given level of environmental variability (reviewed
in Kinzig et al. 2001; Loreau 2010), whereas coexistence studies ask how
species richness responds to different levels of environmental
variability (Chesson \& Warner 1981).

@@

We developed a semi-discrete consumer-resource model that allows many
species to coexist on one resource by either the storage effect or
relative nonlinearity. In our model, the consumer can be in one of
two-states: a dormant state \(D\) and a live state \(N\). The dormant
state could represent, for example, the seed bank of an annual plant
\new{or root biomass of a perennial plant}. Transitions between \(N\)
and \(D\) occur at discrete intervals between growing seasons, with
continuous-time consumer-resource dynamics between the discrete
transitions. Thus, our model is formulated as ``pulsed differential
equations'' (Pachepsky et al. 2008; Mailleret \& Lemesle 2009; Mordecai
et al. 2016).
\new{For clarity we refer to $\tau$ as seasons and the growing time within seasons have daily ($t$) time steps ($t=1,2,3,\dots,T$).
For example, the notation $\tau(t)$ reads as: ``day $t$ within growing season $\tau$.''}

@@

\begin{align}
N_{i,\tau(0)} = \gamma_{i,\tau} D_{i,\tau(0)},
\end{align}\vspace{-3em}

\noindent \new{where \emph{i} indexes each species and $\tau(0)$ denotes the beginning of growing season $\tau$.}
\new{Live biomass at the start of the growing season $\left(N_{i,\tau(0)}\right)$ then serves as the initial conditions for continuous-time}
consumer-resource dynamics that are modeled as two differential
equations: \vspace{-3em}

@@

\new{At the end of the growing season ($t=T$), a fraction ($\alpha_i$) of live biomass is stored as dormant biomass and a fraction of dormant biomass survives ($\eta_i$) to the next growing season, giving the following equation:}
\vspace{-3em}

\begin{align}
D_{i,\tau + 1} = \left[\alpha_i N_{i,\tau(T)} + D_{i,\tau(T)} \right](1 - \eta_i)
\end{align}\vspace{-3em}

\noindent \new{ where $\tau(T)$ denotes the end of growing season $\tau$.
We assume remaining live biomass $\left(N_{\tau(T)}(1-\alpha)\right)$ dies (i.e., this is not a closed system where all biomass must be in either $N$ or $D$ states).}

We assume the resource pool is not replenished within a growing season.
Resource replenishment occurs between growing seasons, and the resource
pool (\emph{R}) at the start of the growing season is
\(R_{\tau(0)} = R^+\), where \(R^+\) is a random resource pulse drawn
from a log-normal distribution with mean \(\mu(R^+)\) and standard
deviation \(\sigma(R^+)\).
\new{Taken all together, we can combine equations 1 and 4 to define the discrete transitions between live and dormant biomass at the end of a growing season.
Thus, the initial conditions for each state ($D, N, R$) at the beginning of growing season $\tau+1$ are:}
\vspace{-3em}

\begin{align}
  D_{i,\tau(0) + 1} &= (1-\gamma_{i,\tau}) \left[\alpha_i N_{i,\tau(T)} + D_{i,\tau(T)} \right] (1-\eta_i) \\
  N_{i,\tau(0) + 1} &= \gamma_{i,\tau} \left[\alpha_i N_{i,\tau(T)} + D_{i,\tau(T)} \right] (1-\eta_i) \\
  R_{\tau + 1} &= \text{lognormal}\left(\mu(R^+), \sigma(R^+) \right),
\end{align}\vspace{-3em}

\noindent \new{where, as above, $\tau(T)$ denotes the end of growing season $\tau$ and $\tau(0) + 1$ denotes the beginning of growing season $\tau+1$.}
Model parameters and notation are described in Table 1.

@@

When we allowed the variance of the environment to determine which of
four initial species coexisted, similar to a study across a natural
diversity gradient, we found a positive relationship between richness
and \new{ecosystem variability}, defined as the temporal \emph{CV} of
total community biomass (Fig. 2A,C). This was true for the storage
effect, where coexistence is maintained by fluctuating dormant-to-live
transition rates (\(\gamma\)), and for relative nonlinearity, where
coexistence is maintained by annual resource pulses. The relationship is
driven by the fact that increasing environmental variability increases
the strength of both coexistence mechanisms (Fig. S1-2). More variable
conditions promoted species richness, creating a positive relationship
between diversity and \new{ecosystem} variability.

When we performed species removals but held environmental variability at
a level that allows coexistence of all four species, similar to a
biodiversity-ecosystem functioning experiment, we found a negative
diversity-\new{ecosystem variability} relationship (Fig. 2B,D). Scatter
around the relationship was small for the storage effect because all
species have similar temporal variances. Regardless of species identity,
the presence of more species always stabilized ecosystem functioning
through portfolio effects. In contrast, scatter around the relationship
was larger for relative nonlinearity (Fig. 2D) because species with
different resource uptake curves had different population variances.
Depending on which species were present, two-species communities were
sometimes less variable than three-species communities.

@@

The dampening effect of fluctuation-dependent coexistence on increasing
environmental variability depends on the specific traits (parameter
values) of the species in the regional pool. Moderate competition makes
it more difficult for new species to enter the local community, but once
they do enter, ecosystem \emph{CV} is similar between communities with
low and moderate competition (Fig. 3; compare top and bottom panels).
Moderate competition does increase the rate at which ecosystem \emph{CV}
increases with environmental variance (Fig. S1-3) because the abundance
of inferior competitors is reduced and they do not influence ecosystem
\emph{CV} as much as when competition is low. The correlation of
species' environmental responses also mediates the relationship between
environmental variance, species richness, and ecosystem \emph{CV}: lower
correlations make it easier for new species to enter the community and
contribute to porfolio effects (Fig. 3).

@@

Theory developed for biodiversity-ecosystem function experiments
emphasizes that increases in species richness should reduce ecosystem
variability. Consistent with theoretical expectations from models and
with results from biodiversity-ecosystem functioning experiments, in
which species coexistence is maintained by fluctuation-independent
mechanisms, our model of fluctuation-dependent species coexistence (also
see Chesson et al. 2001) produced a negative
diversity-\new{ecosystem variability} relationship (Fig. 2B,D). This
agreement is encouraging because species almost certainly coexist by
some combination of fluctuation-independent (e.g., resource
partitioning) and fluctuation-dependent mechanisms. By extending theory
to communities where species richness is explicitly maintained by
temporal variability, we have gained confidence that experimental
findings are generalizable to many communities. In local settings where
environmental variability is relatively homogeneous, reductions in the
number of species should increase the variability of ecosystem
functioning, regardless of how coexistence is maintained.

@@

Our results may explain why deviations from the negative
diversity-\new{ecosystem variability} relationship often come from
observational studies (Jiang \& Pu 2009). Observational studies must
rely on natural diversity gradients, which do not control for
differences in environmental variability among sites. If species
richness depends on environmental variability, it is entirely possible
to observe positive diversity-\new{ecosystem variability} relationships.
For example, Sasaki and Lauenroth (2011) found a positive relationship
between species richness and the temporal variability of plant abundance
in a semi-arid grassland. Their data came from a six sites that were 6
km apart. While Sasaki and Lauenroth explained their results in terms of
dominant species' effects, it is also possible that each site
experienced sufficiently different levels of environmental variability
to influence species coexistence. DeClerck et al. (2006) also found a
positive diversity-\new{ecosystem variability} when sampling conifer
richness and the variability of productivity across a large spatial
gradient in the Sierra Nevada.

While our modeling results show that fluctuation-dependent coexistence
can create positive diversity-\new{ecosystem variability} relationships,
whether such trends are detected will depend on the particular traits of
the species in the community and the relative influence of
fluctuation-dependent and fluctuation-independent coexistence
mechanisms. Thus, our results may also help explain observational
studies where no relationship between diversity and variability is
detected. For example, Cusson et al. (2015) found no relationship
between species richness and variability of abundances in several marine
macro-benthic ecosystems. Many of their focal ecosystems were from
highly variable intertidal environments. If coexistence was at least in
part determined by environmental fluctuations, then the confounding
effect of environmental variability and species richness could offset or
overwhlem any effect of species richness on ecosystem variability.
Previous theoretical work showed how environmental variation can mask
the effect of species diversity on ecosystem productivity when sampling
across sites (Loreau 1998). Our mechanistic model extends that
conclusion to ecosystem variability.
